---
title: "Australian chloropleths in R"
author: "Hugh Parsonage"
date: 2017-06-17
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r loadPackages, message=FALSE}
library(scales)
library(ggplot2)
library(ggthemes)
library(broom)
library(testthat)
library(htmltools)
library(leaflet)
library(viridis)
library(dtplyr)
library(dplyr)
library(magrittr)
library(data.table)
library(jsonlite)
library(rgdal)
library(htmlwidgets)  # saving leaflet maps
```

This post was inspired by Peter Ellis's tutorial: https://ellisp.github.io/blog/2017/06/04/military-gdp.

# Creating maps of Australia
Australia is a particularly challenging country to map.
Australia is very large, highly urbanized, with cities far away from each other.

In R, there are three steps in creating an interactive map of Australia:

1. Install and attach the `ASGS` package.
2. Join your data to the `data` slot of the relevant shapefile
3. Pipe this shapefile to `leaflet` and `addPolygons` to it.

I'll take you through these steps by demonstrating how to plot the population of Australia by SA2. 
Population is not terribly interesting, but the data is easily accessible.

## Accessing the shapefiles using the `ASGS` package
I wrote the `ASGS` package to store the shapefiles from the ABS in package form for convenience and reproducibility.
To install, download and install from Dropbox using the following chunk.[^1]

[^1]: I would appreciate any advice in how better to distribute the package. Please file an issue at https://github.com/HughParsonage/ASGS/

```{r loadASGS}
if (!requireNamespace("ASGS", quietly = TRUE)) {
  message("Attempting install of ASGS (700 MB) from Dropbox. ",
          "This should take some minutes to download.")
  tempf <- tempfile(fileext = ".tar.gz")
  download.file(url = "https://dl.dropbox.com/s/zmggqb1wmmv7mqe/ASGS_0.4.0.tar.gz",
                destfile = tempf)
  install.packages(tempf, repos = NULL, type = "source")
}
library(ASGS)
```


## Joining the data

First, we need the data itself. Naturally, this will ordinarily be a task in itself. But the goal is simple: you need a value for each polygon (each SA2 in this case). 

```{r ERP_by_SA2}
ERP_by_SA2 <- fread("https://raw.githubusercontent.com/HughParsonage/ABS-data/e8d85c749052bf15038971848402cb42ce0d6227/Estimated-Resident-Population-SA2-2016.txt")
ERP_by_SA2[, SA2_MAIN11 := factor(ASGS_2011)]
```

## Static plot methods
As an apostrophe, I'll briefly mention the static plot methods available. 
They're reliable, useful when verifying or debugging, and often surprisingly sufficient.

### Using `plot()`
You can plot shapefiles immediately using the `plot` method.

```{r bare-SA2-chloropleth-base-R}
plot(SA2_2011)
```

And it is fairly straightforward to create a chloropleth by adding colours. The following function simply applies palettes given values. Use `palette_v` 

```{r palette_v}
#' @return Viridis palette of x by quantile (decile by default).
palette_v <- function(x, n = 10) {
  input <- data.table(x = x)
  input[, order := 1:.N]
  input[, decile := ntile(x, n)]
  
  colortbl <- data.table(decile = 1:10,
                         colour = viridis(10))
  
  colortbl[input, on = "decile"] %>%
    setorder(order) %>%
    .[["colour"]]
}

pal_v <- colorNumeric(palette = viridis(10),
                      domain = NULL)
```

```{r SA2_2011_join}
# Copy shapefile and join with ERP_by_SA2 to create
# a column to fill by.
SA2_2011_join <- 
  SA2_2011

SA2_2011_join@data <-
  inner_join(SA2_2011_join@data, ERP_by_SA2, by = "SA2_MAIN11") %>%
  # + 1 in case any 0sqm regions create NaNs
  mutate(Density = Value / ((ALBERS_SQM + 1) / 1e6),
         Density_Decile = ntile(Density, 10))
```

```{r chloropleth-base-R-Population}
plot(SA2_2011_join,
     col = palette_v(SA2_2011_join@data$Value),
     main = "Population by SA2")
```

```{r chloropleth-base-R-PopulationDensity}
plot(SA2_2011_join,
     col = palette_v(SA2_2011_join@data$Density_Decile),
     main = "Population density by SA2")
```

Maps so produced can often be surprisingly 'sufficient'. I often run `dev.copy2pdf(file = "my-map.pdf", width = 200, height = 150)` to 
create a PDF to look closer at the cities. It's not quite as natural or rich in features as mapps produced via `leaflet()` are below, but it's simple, very reliable,
and looks great when you zoom in.

If you're creating a single, static map to be included in a PDF, I would run `dev.copy2pdf` here, pound LaTeX with `trim` and `clip` options to `\includegraphics` for about 15 minutes, then `TikZ` the rest. I doubt what I would produce would be worse than doing it the canonical way.

However, if the statistic for creating the colours in the chloropleth are likely to change, or you need more flexibility, you should use `ggplot2` with the `geom_polygon` layer to create the map.

### Using `ggplot2`
Unlike `plot()` and `leaflet()` you can't just magically pipe shapefiles into `ggplot()`, you must first `broom::tidy` them. Slightly disconcertingly, you need to specify the key using the rownames of the tidied shapefile. However, once you overcome that hurdle, you have the immense power of `ggplot2` to work with.

```{r ggplot-population}
SA2_population_by_id <- ERP_by_SA2[, id := as.character(1:.N - 1)]

SA2_2011 %>%
  tidy %>%
  setDT %>%
  SA2_population_by_id[., on = "id"] %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Value)) + 
  geom_polygon() + 
  scale_fill_viridis("Population", labels = comma) + 
  coord_map() +
  theme_map()
```

Nonetheless, the urban areas are hidden, and rural areas are ridiculously overemphasized.

## Interactive plots using leaflet
Because of the challenges of mapping in Australia,
interactive chloropleths have considerable appeal.
Leaflet through its eponymous R package offers this feature.

Like with `plot`, `leaflet()` accepts shapefiles as inputs, and you can pass shapefiles immediately.
However, for the SA2 shapefiles, this results in hefty HTML files. 
On my machine, RStudio creaked under their weight.

(You should run the following two chunks. I've set `eval=FALSE` to reduce the file size of this page from 250 MB to 25 MB!)

```{r leaflet-Population, eval=FALSE}
SA2_2011_join %>%
  leaflet %>%
  addPolygons
```

The default settings of leaflet are not suitable for presentation; however, leaflet offers enough flexibility to create elegant maps:

```{r leaflet-Population-viridis, eval=FALSE}
SA2_2011_join %>%
  leaflet %>%
  addPolygons(stroke = TRUE,
              opacity = 1,
              weight = 1,  # the border thickness / pixels
              color = "white",
              fillColor = ~pal_v(SA2_2011_join@data$Density_Decile), 
              fillOpacity = 1)
```

However, the file is too large (125 MB on my computer) to be shared easily.  `library(ASGS)` offers `SA2_2011_simple`, a simplified version (20\%) of `SA2_2011` produced from the GUI at http://mapshaper.org/.

Further compression of the HTML file can be achieved by reducing the precision of the coordinates.
In contrast to simplifying the shapefile (which reduces the detail of borders), reducing the precision does not change the shape of borders much, but translates them east or north.

The ABS's standard uses a nominal precision of about 1 mm, which is vastly more precise than needed for SA2 boundaries. 
By reducing the precision, the raw HTML (which lists the coordinates of every vertex) will have literally use fewer characters to mark each coordinate, shaving off a considerable chunk of the size of your page, improving performance.

I chose a precision of 0.0001 which means the borders can be translated at most 10 metres. Such a modest loss of precision, even for detailed places like inner Sydney and Melbourne, had no visible effect.

The file in the package was created simply  by running (again in http://mapshaper.org/)

```bash
mapshaper -o precision=0.0001 format=topojson out.json
```

The resulting chloropleth is 15 MB, which I suspect cannot
be improved. (Reducing the precision further resulted in unsightly overlaps at Walsh Bay.)

```{r signif_comma}
#' @param x
#' @param digits
#' @return For values over 1000, a comma()'d version of x to \code{digits} significant figures;  otherwise, 
#' x to \code{digits} significant places.
#' 
#' Using \code{comma(signif(x, 2))} would produce '120,000.0' instead of '120,000'.
signif_comma <- function(x, digits = 2) {
  y <- signif(x, digits = digits)
  out <- y
  out[log10(x) > 3] <- scales::comma(y[log10(x) > 3])
  out
}

expect_equal(signif_comma(c(1.2, 12, 1200, 12000, 120000)),
             c("1.2", "12", "1,200", "12,000", "120,000"))
```

```{r SA2_2011_simple_-chloropleth-leaflet, fig.width=10}
# Copy shapefile and join with ERP_by_SA2 to create
# a column to fill by.

SA2_2011_simple %>%
  leaflet %>%
  addPolygons(stroke = TRUE,
              opacity = 1,
              weight = 1,  # the border thickness / pixels
              color = "black",
              fillColor = ~pal_v(SA2_2011_join@data$Density_Decile), 
              fillOpacity = 1,
              label = lapply(paste0("<b>", 
                                    SA2_2011_join@data$SA2_NAME11, ":",
                                    "</b><br>",
                                    signif_comma(SA2_2011_join@data$Density, 2),
                                    "/km²"),
                             HTML),
              highlightOptions = highlightOptions(weight = 2,
                                                  color = "white",
                                                  opacity = 1,
                                                  dashArray = "",
                                                  bringToFront = TRUE))
```

This provides a lightweight, interactive chloropleth, as required. 




## Just give me the code

```{r, united, eval=FALSE}
## ----setup, include=FALSE------------------------------------------------
knitr::opts_chunk$set(collapse = TRUE)

## ----loadPackages, message=FALSE-----------------------------------------
library(scales)
library(ggplot2)
library(ggthemes)
library(broom)
library(testthat)
library(htmltools)
library(leaflet)
library(viridis)
library(dtplyr)
library(dplyr)
library(magrittr)
library(data.table)
library(jsonlite)
library(rgdal)
library(htmlwidgets)  # saving leaflet maps

## ----loadASGS------------------------------------------------------------
if (!requireNamespace("ASGS", quietly = TRUE)) {
  message("Attempting install of ASGS (500 MB) from Dropbox. ",
          "This should take some minutes to download.")
  tempf <- tempfile(fileext = ".zip")
  download.file(url = "https://www.dropbox.com/s/acn5v30yco4jrx7/ASGS_0.0.3.zip?dl=1",
                destfile = tempf)
  install.packages(tempf, repos = NULL)
}
library(ASGS)

## ----ERP_by_SA2----------------------------------------------------------
ERP_by_SA2 <- fread("https://raw.githubusercontent.com/HughParsonage/ABS-data/master/Estimated-Resident-Population-SA2-2016.txt")
ERP_by_SA2[, SA2_MAIN11 := factor(ASGS_2011)]

## ----bare-SA2-chloropleth-base-R-----------------------------------------
plot(SA2_2011)

## ----palette_v-----------------------------------------------------------
#' @return Viridis palette of x by quantile (decile by default).
palette_v <- function(x, n = 10) {
  input <- data.table(x = x)
  input[, order := 1:.N]
  input[, decile := ntile(x, n)]
  
  colortbl <- data.table(decile = 1:10,
                         colour = viridis(10))
  
  colortbl[input, on = "decile"] %>%
    setorder(order) %>%
    .[["colour"]]
}

pal_v <- colorNumeric(palette = viridis(10),
                      domain = NULL)

## ----SA2_2011_join-------------------------------------------------------
# Copy shapefile and join with ERP_by_SA2 to create
# a column to fill by.
SA2_2011_join <- 
  SA2_2011

SA2_2011_join@data <-
  inner_join(SA2_2011_join@data, ERP_by_SA2, by = "SA2_MAIN11") %>%
  # + 1 in case any 0sqm regions create NaNs
  mutate(Density = Value / ((ALBERS_SQM + 1) / 1e6),
         Density_Decile = ntile(Density, 10))

## ----chloropleth-base-R-Population---------------------------------------
plot(SA2_2011_join,
     col = palette_v(SA2_2011_join@data$Value),
     main = "Population by SA2")

## ----chloropleth-base-R-PopulationDensity--------------------------------
plot(SA2_2011_join,
     col = palette_v(SA2_2011_join@data$Density_Decile),
     main = "Population density by SA2")

## ----ggplot-population---------------------------------------------------
SA2_population_by_id <- ERP_by_SA2[, id := as.character(1:.N - 1)]

SA2_2011 %>%
  tidy %>%
  setDT %>%
  SA2_population_by_id[., on = "id"] %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Value)) + 
  geom_polygon() + 
  scale_fill_viridis("Population", labels = comma) + 
  coord_map() +
  theme_map()

## ----leaflet-Population, eval=FALSE--------------------------------------
## SA2_2011_join %>%
##   leaflet %>%
##   addPolygons

## ----leaflet-Population-viridis, eval=FALSE------------------------------
## SA2_2011_join %>%
##   leaflet %>%
##   addPolygons(stroke = TRUE,
##               opacity = 1,
##               weight = 1,  # the border thickness / pixels
##               color = "white",
##               fillColor = ~pal_v(SA2_2011_join@data$Density_Decile),
##               fillOpacity = 1)

## ----signif_comma--------------------------------------------------------
#' @param x
#' @param digits
#' @return For values over 1000, a comma()'d version of x to \code{digits} significant figures;  otherwise, 
#' x to \code{digits} significant places.
#' 
#' Using \code{comma(signif(x, 2))} would produce '120,000.0' instead of '120,000'.
signif_comma <- function(x, digits = 2) {
  y <- signif(x, digits = digits)
  out <- y
  out[log10(x) > 3] <- scales::comma(y[log10(x) > 3])
  out
}

expect_equal(signif_comma(c(1.2, 12, 1200, 12000, 120000)),
             c("1.2", "12", "1,200", "12,000", "120,000"))

## ----SA2_2011_simple_-chloropleth-leaflet, fig.width=10------------------
# Copy shapefile and join with ERP_by_SA2 to create
# a column to fill by.

SA2_2011_simple %>%
  leaflet %>%
  addPolygons(stroke = TRUE,
              opacity = 1,
              weight = 1,  # the border thickness / pixels
              color = "black",
              fillColor = ~pal_v(SA2_2011_join@data$Density_Decile), 
              fillOpacity = 1,
              label = lapply(paste0("<b>", 
                                    SA2_2011_join@data$SA2_NAME11, ":",
                                    "</b><br>",
                                    signif_comma(ERP_by_SA2$, 2),
                                    "/km²"),
                             HTML),
              highlightOptions = highlightOptions(weight = 2,
                                                  color = "white",
                                                  opacity = 1,
                                                  dashArray = "",
                                                  bringToFront = TRUE))
## ----smoothed-chloropleth, fig.width=10
y <- smooth_shapefile(shapefile = SA2_2011,
                      data_by_id = ERP_by_SA2,
                      id = "SA2_MAIN11",
                      var2smooth = "Value",
                      coalesce.to = as.integer(mean(ERP_by_SA2$Value, na.rm = TRUE)), 
                      k = 10)

pal_v <- colorNumeric(palette = viridis::viridis(6), domain = NULL)

SA2_2011_simple %>%
  leaflet() %>%
  addPolygons(fillOpacity = 1,
              stroke = TRUE,
              weight = 1,
              color = "black",
              fillColor = ~pal_v(y),
              opacity = 1,
              smoothFactor = 2,
              label = lapply(paste0("<strong>",
                                    as.character(SA2_2011@data$SA2_NAME11), ":",
                                    "</strong><br>",
                                    signif_comma(as.numeric(ERP_by_SA2$Value))),
                             HTML),
              highlightOptions = highlightOptions(weight = 2,
                                                  color = "white",
                                                  opacity = 1,
                                                  dashArray = "",
                                                  bringToFront = TRUE))


```
