---
title: "A new ifelse"
author: "Hugh Parsonage"
date: "3 May 2018"
output: html_document
---

```{r knitOpts}
library(knitr)
options(digits = 2)
```

## Summary

`ifelse` can be made significantly faster without loss of backwards-compatibility by using `anyNA` and falling back to the original code whenever backwards-compatibility cannot be assured.


### Current `ifelse`:

```{r new_ifelse-original}
old_ifelse <- function (test, yes, no)
{
    if(is.atomic(test)) { # do not lose attributes
        if (typeof(test) != "logical")
            storage.mode(test) <- "logical"
        ## quick return for cases where 'new_ifelse(a, x, y)' is used
        ## instead of 'if (a) x else y'
        if (length(test) == 1 && is.null(attributes(test))) {
            if (is.na(test)) return(NA)
            else if (test) {
                if (length(yes) == 1) {
                    yat <- attributes(yes)
                    if (is.null(yat) || (is.function(yes) &&
                                         identical(names(yat), "srcref")))
                        return(yes)
                }
            }
            else if (length(no) == 1) {
                nat <- attributes(no)
                if (is.null(nat) || (is.function(no) &&
                                     identical(names(nat), "srcref")))
                    return(no)
            }
        }
    }
    else ## typically a "class"; storage.mode<-() typically fails
	test <- if(isS4(test)) methods::as(test, "logical") else as.logical(test)
    ans <- test
    ok <- !is.na(test)
    if (any(test[ok]))
	ans[test & ok] <- rep(yes, length.out = length(ans))[test & ok]
    if (any(!test[ok]))
	ans[!test & ok] <- rep(no, length.out = length(ans))[!test & ok]
    ans
}
```

The proposal is to change `ifelse` to take advantage of the speed of `anyNA()`, use `is.na` only once after (and only if needed), and to take advantage of the predictable nature of `ifelse` when the lengths of `test`, `yes`, and `no` are equal or length-one.

I also correct an apparent bug in `ifelse` where the attributes of `test` are not saved if `isS4(test)`. 

```{r new_ifelse}
.new_ifelse <- function(test, yes, no) {
  ans <- test
  ok <- !is.na(test)
  if (any(test[ok]))
    ans[test & ok] <- rep(yes, length.out = length(ans))[test & ok]
  if (any(!test[ok]))
    ans[!test & ok] <- rep(no, length.out = length(ans))[!test & ok]
  ans
}

new_ifelse <- function (test, yes, no)
{
  attributes_of_test <- attributes(test)
  
  if(is.atomic(test)) { # do not lose attributes
    if (typeof(test) != "logical")
      storage.mode(test) <- "logical"
    ## quick return for cases where 'new_ifelse(a, x, y)' is used
    ## instead of 'if (a) x else y'
    if (length(test) == 1 && is.null(attributes_of_test)) {
      if (is.na(test)) return(NA)
      else if (test) {
        if (length(yes) == 1) {
          yat <- attributes(yes)
          if (is.null(yat) || (is.function(yes) &&
                               identical(names(yat), "srcref")))
            return(yes)
        }
      }
      else if (length(no) == 1) {
        nat <- attributes(no)
        if (is.null(nat) || (is.function(no) &&
                             identical(names(nat), "srcref")))
          return(no)
      }
    }
  }
  else ## typically a "class"; storage.mode<-() typically fails
    test <- if(isS4(test)) methods::as(test, "logical") else as.logical(test)
  
  # Give up attempting backwards-compatibility under these conditions:
  if (typeof(yes) %in% c("logical", "integer", "double", "character") &&
      typeof(no)  %in% c("logical", "integer", "double", "character") &&
      !is.factor(yes) &&
      !is.factor(no) &&
      length(no) != 0L &&
      length(yes) != 0L) {
    if (length(no) == length(test)) {
      out <- no
    } else if (length(no) == 1L) {
      out <- rep_len(no, length(test))
    } else {
      return(.new_ifelse(test, yes, no))
    }
    
    if (length(yes) != 1L && length(yes) != length(test)) {
      return(.new_ifelse(test, yes, no))
    }
    
    
    if (anyNA(test)) {
      # no benefit to saving the na results
      Yes <- which(test)
      out[is.na(test)] <- NA
      if (length(yes) == 1L) {
        out[Yes] <- yes
      } else if (length(yes) == length(test)) {
        out[Yes] <- yes[Yes]
      } else {
        return(.new_ifelse(test, yes, no))
      }
    } else {
      # No NAs to deal with
      if (length(yes) == 1L) {
        out[test] <- yes
      } else if (length(yes) == length(test)) {
        wtest <- which(test) # faster than test directly
        out[wtest] <- yes[wtest]
      } else {
        return(.new_ifelse(test, yes, no))
      }
    }
    if (!is.null(attributes_of_test)) {
      attributes(out) <- attributes_of_test
    }
    
    out
  } else {
    ans <- test
    ok <- !is.na(test)
    if (any(test[ok]))
      ans[test & ok] <- rep(yes, length.out = length(ans))[test & ok]
    if (any(!test[ok]))
      ans[!test & ok] <- rep(no, length.out = length(ans))[!test & ok]
    ans
  }
}

new_ifelse100 <- function (test, yes, no)
{
  attributes_of_test <- attributes(test)
  
  if(is.atomic(test)) { # do not lose attributes
    if (typeof(test) != "logical")
      storage.mode(test) <- "logical"
    ## quick return for cases where 'new_ifelse(a, x, y)' is used
    ## instead of 'if (a) x else y'
    if (length(test) == 1 && is.null(attributes_of_test)) {
      if (is.na(test)) return(NA)
      else if (test) {
        if (length(yes) == 1) {
          yat <- attributes(yes)
          if (is.null(yat) || (is.function(yes) &&
                               identical(names(yat), "srcref")))
            return(yes)
        }
      }
      else if (length(no) == 1) {
        nat <- attributes(no)
        if (is.null(nat) || (is.function(no) &&
                             identical(names(nat), "srcref")))
          return(no)
      }
    }
  }
  else ## typically a "class"; storage.mode<-() typically fails
    test <- if(isS4(test)) methods::as(test, "logical") else as.logical(test)
  
  # Give up attempting backwards-compatibility under these conditions:
  if (length(test) > 100L &&
      typeof(yes) %in% c("logical", "integer", "double", "character") &&
      typeof(no)  %in% c("logical", "integer", "double", "character") &&
      !is.factor(yes) &&
      !is.factor(no) &&
      length(no) != 0L &&
      length(yes) != 0L) {
    if (length(no) == length(test)) {
      out <- no
    } else if (length(no) == 1L) {
      out <- rep_len(no, length(test))
    } else {
      return(.new_ifelse(test, yes, no))
    }
    
    if (length(yes) != 1L && length(yes) != length(test)) {
      return(.new_ifelse(test, yes, no))
    }
    
    
    if (anyNA(test)) {
      # no benefit to saving the na results
      Yes <- which(test)
      out[is.na(test)] <- NA
      if (length(yes) == 1L) {
        out[Yes] <- yes
      } else if (length(yes) == length(test)) {
        out[Yes] <- yes[Yes]
      } else {
        return(.new_ifelse(test, yes, no))
      }
    } else {
      # No NAs to deal with
      if (length(yes) == 1L) {
        out[test] <- yes
      } else if (length(yes) == length(test)) {
        wtest <- which(test) # faster than test directly
        out[wtest] <- yes[wtest]
      } else {
        return(.new_ifelse(test, yes, no))
      }
    }
    if (!is.null(attributes_of_test)) {
      attributes(out) <- attributes_of_test
    }
    
    out
  } else {
    ans <- test
    ok <- !is.na(test)
    if (any(test[ok]))
      ans[test & ok] <- rep(yes, length.out = length(ans))[test & ok]
    if (any(!test[ok]))
      ans[!test & ok] <- rep(no, length.out = length(ans))[!test & ok]
    ans
  }
}

```


### Unit tests

```{r test_identicals}
test_identicals <- function(test, yes, no) {
  old <- old_ifelse(test, yes, no)
  new <- new_ifelse(test, yes, no)
  if (!identical(old, new)) {
    # Will take a few minutes otherwise
    if (length(old)  + length(new) < 50) {
      print(old)
      print(new)
    }
    cat("Non-identical result\n")
    return(invisible(list(old = old, 
                          new = new)))
  }
}
```

```{r unit-tests}
# Attributes?
test1 <- c(a = FALSE, b = FALSE, c = TRUE, d = TRUE, e = TRUE)
test_identicals(test1, 1:4 + 0.0, 3)
yes1 <- 1:4 + 0
names(yes1) <- "srcref"
test_identicals(test1, yes1, 2:3)

test2 <- sample(test1, size = 1e7, replace = TRUE)
test_identicals(test2, 1:4 + 0.0, 3)
test2[5e5] <- NA
test_identicals(test2, 1:4 + 0.0, 3)
```

As aformentioned, `S4` classes did not retain attributes:
```{r}
if (requireNamespace("Matrix", quietly = TRUE)) {
  library(Matrix)
  M <- Matrix(-10 + 1:28, 4, 7)
  test_identicals(M, 1, 2)
}
```

These examples simply fall back to the defaults.

```{r}
test_identicals(c(TRUE, TRUE), "a", raw(0x95))
test_identicals(c(TRUE, TRUE), "a", factor("a"))
test_identicals(c(FALSE, FALSE), "a", factor("a"))
test_identicals(c(FALSE, TRUE), 1 , factor("a"))
test_identicals(c(FALSE, TRUE, NA), 1, factor("a"))
```


## Benchmarks

```{r small-test-single-yes-no, fig.width=11, fig.height=11}
library(data.table)
library(ggplot2)
library(microbenchmark)
library(scales)
get_avg_time <- function(n, yes1 = TRUE, include_NA = FALSE) {
  test <- sample(c(TRUE, FALSE, if (include_NA) NA), size = n, replace = TRUE)
  if (yes1) {
    yes <- -1L
  } else {
    yes <- sample.int(n)
  }
  no <- 1L
  old <- as.data.table(microbenchmark(old_ifelse(test, yes, no)), times = if (n < 1e3) 100 else 10)
  new <- as.data.table(microbenchmark(new_ifelse(test, yes, no)), times = if (n < 1e3) 100 else 10)
  new100 <- as.data.table(microbenchmark(new_ifelse100(test, yes, no)), times = if (n < 1e3) 100 else 10)
  
  nmed <- new100[, median(time)]
  omed <- old[, median(time)]
  list(old[, median(time)], 
       new[, median(time)],
       new100[, median(time)], 
       prop_faster = abs(new100[, mean(time < omed)] - old[, mean(time < nmed)]))
}

DT <- 
  CJ(n = c(1:10, 10*(2:10), 50*(3:10), 750, 1000*(1:10), 10e3, 50e3, 100e3),
     yes1 = c(TRUE, FALSE), 
     include_NA = c(TRUE, FALSE))
DT[, I := .I]
DT[, c("old", "new", "new100", "prop_faster") := get_avg_time(n, yes1, include_NA), by = "I"]
DT[, f1 := "without NAs"]
DT[(include_NA), f1 := "with NAs"]
DT[, f2 := "yes length-n"]
DT[(yes1), f2 := "yes length-1"]
ggplot(melt(DT, measure.vars = c("old", "new", "new100"))[, `time (ms)` := value / 1e6],
       aes(x = n, y = `time (ms)`, color = variable, alpha = prop_faster)) + 
  geom_line(size = 1.1) + 
  facet_wrap(~ f2 + f1) + 
  scale_x_log10(labels = scales::comma, breaks = c(1, 10, 100, 1000, 100e3)) + 
  scale_y_log10(labels = scales::comma)
```




