---
title: "A new ifelse"
author: "Hugh Parsonage"
date: "3 May 2018"
output: html_document
---

```{r knitOpts}
library(knitr)
options(digits = 2)
```

## Summary

`ifelse` can be made significantly faster without loss of backwards-compatibility by using `anyNA` and falling back to the original code


### Current `ifelse`:

```{r ifelse-original}
ifelse <- function (test, yes, no)
{
    if(is.atomic(test)) { # do not lose attributes
        if (typeof(test) != "logical")
            storage.mode(test) <- "logical"
        ## quick return for cases where 'ifelse(a, x, y)' is used
        ## instead of 'if (a) x else y'
        if (length(test) == 1 && is.null(attributes(test))) {
            if (is.na(test)) return(NA)
            else if (test) {
                if (length(yes) == 1) {
                    yat <- attributes(yes)
                    if (is.null(yat) || (is.function(yes) &&
                                         identical(names(yat), "srcref")))
                        return(yes)
                }
            }
            else if (length(no) == 1) {
                nat <- attributes(no)
                if (is.null(nat) || (is.function(no) &&
                                     identical(names(nat), "srcref")))
                    return(no)
            }
        }
    }
    else ## typically a "class"; storage.mode<-() typically fails
	test <- if(isS4(test)) methods::as(test, "logical") else as.logical(test)
    ans <- test
    ok <- !is.na(test)
    if (any(test[ok]))
	ans[test & ok] <- rep(yes, length.out = length(ans))[test & ok]
    if (any(!test[ok]))
	ans[!test & ok] <- rep(no, length.out = length(ans))[!test & ok]
    ans
}
```

The proposal is to change `ifelse` to take advantage of the speed of `anyNA()`, use `is.na` only once after (and only if needed), and to take advantage of the predictable nature of `ifelse` when the lengths of `test`, `yes`, and `no` are equal or length-one.

I also correct an apparent bug in `ifelse` where the attributes of `test` are not saved if `isS4(test)`. 

```{r new-ifelse}
.ifelse <- function(test, yes, no) {
  ans <- test
  ok <- !is.na(test)
  if (any(test[ok]))
    ans[test & ok] <- rep(yes, length.out = length(ans))[test & ok]
  if (any(!test[ok]))
    ans[!test & ok] <- rep(no, length.out = length(ans))[!test & ok]
  ans
}

ifelse <- function (test, yes, no)
{
  attributes_of_test <- attributes(test)
  
  if(is.atomic(test)) { # do not lose attributes
    if (typeof(test) != "logical")
      storage.mode(test) <- "logical"
    ## quick return for cases where 'ifelse(a, x, y)' is used
    ## instead of 'if (a) x else y'
    if (length(test) == 1 && is.null(attributes_of_test)) {
      if (is.na(test)) return(NA)
      else if (test) {
        if (length(yes) == 1) {
          yat <- attributes(yes)
          if (is.null(yat) || (is.function(yes) &&
                               identical(names(yat), "srcref")))
            return(yes)
        }
      }
      else if (length(no) == 1) {
        nat <- attributes(no)
        if (is.null(nat) || (is.function(no) &&
                             identical(names(nat), "srcref")))
          return(no)
      }
    }
  }
  else ## typically a "class"; storage.mode<-() typically fails
    test <- if(isS4(test)) methods::as(test, "logical") else as.logical(test)
  
  # Give up attempting backwards-compatibility under these conditions:
  if (typeof(yes) %in% c("logical", "integer", "double", "character") &&
      typeof(no)  %in% c("logical", "integer", "double", "character") &&
      !is.factor(yes) &&
      !is.factor(no) &&
      length(no) != 0L &&
      length(yes) != 0L) {
    if (length(no) == length(test)) {
      out <- no
    } else if (length(no) == 1L) {
      out <- rep_len(no, length(test))
    } else {
      return(.ifelse(test, yes, no))
    }
    
    if (length(yes) != 1L && length(yes) != length(test)) {
      return(.ifelse(test, yes, no))
    }
    
    
    if (anyNA(test)) {
      # no benefit to saving the na results
      Yes <- which(test)
      out[is.na(test)] <- NA
      if (length(yes) == 1L) {
        out[Yes] <- yes
      } else if (length(yes) == length(test)) {
        out[Yes] <- yes[Yes]
      } else {
        return(.ifelse(test, yes, no))
      }
    } else {
      # No NAs to deal with
      if (length(yes) == 1L) {
        out[test] <- yes
      } else if (length(yes) == length(test)) {
        wtest <- which(test) # faster than test directly
        out[wtest] <- yes[wtest]
      } else {
        return(.ifelse(test, yes, no))
      }
    }
    if (!is.null(attributes_of_test)) {
      attributes(out) <- attributes_of_test
    }
    
    out
  } else {
    return(.ifelse(test, yes, no))
  }
}
```


### Unit tests

```{r test_identicals}
test_identicals <- function(test, yes, no) {
  old <- base::ifelse(test, yes, no)
  new <- ifelse(test, yes, no)
  if (!identical(old, new)) {
    # Will take a few minutes otherwise
    if (length(old)  + length(new) < 50) {
      print(old)
      print(new)
    }
    cat("Non-identical result\n")
    return(invisible(list(old = old, 
                          new = new)))
  }
}
```

```{r unit-tests}
# Attributes?
test1 <- c(a = FALSE, b = FALSE, c = TRUE, d = TRUE, e = TRUE)
test_identicals(test1, 1:4 + 0.0, 3)
yes1 <- 1:4 + 0
names(yes1) <- "srcref"
test_identicals(test1, yes1, 2:3)

test2 <- sample(test1, size = 1e7, replace = TRUE)
test_identicals(test2, 1:4 + 0.0, 3)
test2[5e5] <- NA
test_identicals(test2, 1:4 + 0.0, 3)
```

As aformentioned, `S4` classes did not retain attributes:
```{r}
if (requireNamespace("Matrix", quietly = TRUE)) {
  library(Matrix)
  M <- Matrix(-10 + 1:28, 4, 7)
  test_identicals(M, 1, 2)
}
```

These examples simply fall back to the defaults.

```{r}
test_identicals(c(TRUE, TRUE), "a", raw(0x95))
test_identicals(c(TRUE, TRUE), "a", factor("a"))
test_identicals(c(FALSE, FALSE), "a", factor("a"))
test_identicals(c(FALSE, TRUE), 1 , factor("a"))
test_identicals(c(FALSE, TRUE, NA), 1, factor("a"))
```


## Benchmarks
```{r}
library(microbenchmark)
library(ggplot2)
print_mb <- function(mb) {
  print(mb)
  print(autoplot(mb))
}
```

```{r single-yes-no, fig.cap="No NAs, len(yes)=len(no)=1"}
test1 <- sample(c(TRUE, FALSE), size = 1e6, replace = TRUE)
yes1 <- 1L
no1 <- -1L
print_mb(microbenchmark(ifelse(test1, yes1, no1), base::ifelse(test1, yes1, no1)))
```

```{r multi-yes-no, fig.cap="No NAs, len(yes)=len(no)=len(test)"}
test1 <- sample(c(TRUE, FALSE), size = 1e6, replace = TRUE)
yes1 <- 1:1e6
no1 <- -yes1
print_mb(microbenchmark(ifelse(test1, yes1, no1), base::ifelse(test1, yes1, no1)))
```

```{r multi-yes-no-NAs, fig.cap="NAs, len(yes)=len(no)=len(test)"}
test3 <- test1
test3[5e5] <- NA
print_mb(microbenchmark(ifelse(test3, yes1, no1), base::ifelse(test3, yes1, no1)))
```

```{r small-test-single-yes-no}
for (pow in 1:4) {
  test1 <- sample(c(TRUE, FALSE), size = 10^pow, replace = TRUE)
  yes1 <- 1L
  no1 <- -1L
  print(prettyNum(10^pow, big.mark = ","))
  print_mb(microbenchmark(ifelse(test1, yes1, no1), base::ifelse(test1, yes1, no1)))
}
```




